---
title: "Test of KF algorithms on a linear system"
author: "Aaron A. King, Masoud Mirmomeni"
date: "19 February 2016"
output: 
  html_document:
    toc: true
---

```{r knitr-opts,include=FALSE,purl=FALSE}
require(knitr)
opts_chunk$set(cache=TRUE,cache.extra=rand_seed)
set.seed(1659125922)
```

```{r prelims,message=FALSE,cache=FALSE}
require(ggplot2)
require(scales)
theme_set(theme_bw())
require(reshape2)
require(plyr)
require(magrittr)
require(mvtnorm)
require(pomp)
```

## The linear model

Observation times:
```{r timespan}
t <- seq(1,100)
```


The observation matrix is $C\in \mathbb{R}^{n\times m}$.

```{r obs-matrix}
C <- matrix(c(0, 1,  0, 0,
              0, 1, -1, 0),
            2,4,byrow=TRUE); 
dimX <- ncol(C)
dimY <- nrow(C)
rownames(C) <- paste0("y",seq_len(dimY))
colnames(C) <- paste0("x",seq_len(dimX))
C
```

The state transition matrix is $A\in\mathbb{R}^{m\times m}$.
```{r proc-mean}
# A <- matrix(c(-0.677822, -0.169411,  0.420662,  0.523571,
#                2.87451,  -0.323604, -0.489533, -0.806087,
#               -1.36617,  -0.592326,  0.567114,  0.345142,
#               -0.807978, -0.163305,  0.668037,  0.468286),
#             nrow=dimX,ncol=dimX,byrow=TRUE,
#             dimnames=list(colnames(C),colnames(C)))
A <- matrix(c(-0.677822, -0.169411,  0.420662, 0,
               2.87451,  -0.323604, -0.489533, 0,
              -1.36617,  -0.592326,  0.567114, 0,
              -0.807978, -0.163305,  0.668037, 0.468286),
            nrow=dimX,ncol=dimX,byrow=TRUE,
            dimnames=list(colnames(C),colnames(C)))
A
```

The process noise covariance matrix, $Q\in\mathbb{R}^{m\times m}$.
```{r proc-noise}
Q <- 10*diag(1,dimX)
dimnames(Q) <- dimnames(A)
Q
```

The observation noise covariance matrix, $R\in\mathbb{R}^{n\times n}$.
```{r obs-noise}
R <- 1*matrix(c(1,0.5,0.5,3),nrow=dimY)
dimnames(R) <- list(rownames(C),rownames(C))
R
```

The initial state, $X_0\in\mathbb{R}^{m}$.
```{r x0}
X0 <- setNames(rnorm(4),colnames(C))
X0
```

Now to simulate some data.
```{r generate-data}
N <- length(t)

x <- array(dim=c(dimX,N),dimnames=list(vbl=colnames(C),time=t))
y <- array(dim=c(dimY,N),dimnames=list(vbl=rownames(C),time=t))

xx <- X0
sqrtQ <- chol(Q)
sqrtR <- chol(R)
for (k in seq_along(t)) {
  x[,k] <- xx <- A %*% xx + sqrtQ %*% rnorm(n=dimX)
  y[,k] <- C %*% xx + sqrtR %*% rnorm(n=dimY)
}
```

Plot the data and true states:
```{r plot-data}
melt(list(obs=y,states=x)) %>%
  ggplot(aes(x=time,y=value,group=vbl))+
  geom_line()+
  facet_grid(vbl~.,scales='free_y')+
  theme(legend.position="top")+
  labs(color=NULL)
```

## Kalman filter (KF)

Prediction mean: $M_t=A\,m_t$  
Prediction variance: $V_t=A\,v_t\,A^T+Q$  
Forecast error: $r_t=y_t-C\,M_t$  
Forecast error variance: $W_t=C\,V_t\,C^T+R$  
Filter variance: $v_t=(V_t^{-1}+C^T\,R^{-1}\,C)^{-1}$  
Kalman gain: $K_t=v_t\,C^T\,R^{-1}$  
Filter mean: $m_t=M_t+K_t\,r_t$  

```{r define-kf}
kalmanFilter <- function (y, X0, A, Q, C, R) {
  N <- ncol(y)
  dimX <- length(X0)
  filterMeans <- array(dim=c(dimX,N),dimnames=dimnames(x))
  predMeans <- array(dim=c(dimX,N),dimnames=dimnames(x))
  forecast <- array(dim=c(dimY,N),dimnames=dimnames(y))
  condlogLik <- numeric(N)
  ri <- solve(R)
  cric <- crossprod(C,ri)%*%C
  fm <- X0
  fv <- matrix(0,dimX,dimX)
  for (k in seq_along(t)) {
    predMeans[,k] <- pm <- A%*%fm      # prediction mean
    pv <- A%*%tcrossprod(fv,A)+Q       # prediction variance    
    svdV <- svd(pv,nv=0)
    resid <- y[,k]-C%*%pm              # forecast error
    w <- tcrossprod(C%*%pv,C)+R        # forecast variance
    svdW <- svd(w,nv=0)
    condlogLik[k] <- sum(dnorm(x=crossprod(svdW$u,resid),mean=0,sd=sqrt(svdW$d),log=TRUE))
    pvi <- svdV$u%*%(t(svdV$u)/svdV$d) # prediction precision
    fvi <- pvi+cric                    # filter precision
    svdv <- svd(fvi,nv=0)
    fv <- svdv$u%*%(t(svdv$u)/svdv$d)  # filter variance
    K <- fv%*%crossprod(C,ri)          # Kalman gain
    filterMeans[,k] <- fm <- pm+K%*%resid  # filter mean
    forecast[,k] <- C %*% pm
  }
  list(filterMeans=filterMeans,predMeans=predMeans,forecast=forecast,condlogLik=condlogLik)
}
```

```{r run-kf}
kf <- kalmanFilter(y,X0,A,Q,C,R)
```

Plot the results:
```{r plot-kf,warning=FALSE}
melt(list(obs=list(true=y,pred=kf$forecast),
          states=list(true=x,filter=kf$filterMeans,pred=kf$predMeans))) %>%
  dcast(vbl+time~L2) %>%
  ggplot(aes(x=time))+
  geom_point(aes(y=true),color='black')+
  geom_point(aes(y=pred),color='red',alpha=0.5)+
  geom_line(aes(y=filter),color='blue',alpha=0.3)+
  facet_grid(vbl~.,scales='free_y')+
  theme(legend.position="top")
```

## Particle filter (PF)

```{r pfilter}
require(pomp)
require(mvtnorm)
y %>% melt() %>% dcast(time~vbl) %>%
  pomp(times='time',t0=0,
       rprocess=discrete.time.sim(
         step.fun=function(x,t,params,...){
           A%*%x+sqrtQ%*%rnorm(n=dimX)
         },
         delta.t=1),
       rmeasure=function(x,t,params,...){
         C%*%x+sqrtR%*%rnorm(n=dimY)
       },
       dmeasure=function(y,x,t,params,log,...){
         dmvnorm(x=t(y-C%*%x),sigma=R,log=log)
       },
       initializer=function(params,t0,...){
         X0
       },
       params=c(dummy=3)) %>%
  pfilter(Np=2000,filter.mean=TRUE) -> pf
```

## Ensemble Kalman Filter (EnKF)

Ensemble: $X_t\in \mathbb{R}^{m\times q}$  
Prediction mean: $M_t=\langle X \rangle$  
Prediction variance: $V_t=\langle\langle X \rangle\rangle$  
Forecast: $Y_t=C\,X_t$  
Forecast mean: $N_t=\langle Y \rangle$.  
Forecast variance: $S_t=\langle\langle Y \rangle\rangle$  
State/forecast covariance: $W_t=\langle\langle X,Y\rangle\rangle$  
Kalman gain: $K_t = W_t\,S_t^{-1}$  
New observation: $y_t\in \mathbb{R}^{n\times 1}$  
Updated ensemble: $X^u_{t}=X_t + K_t\,(O_t - Y_t)$  
Filter mean: $m_t=\langle X^u_t \rangle = \frac{1}{q} \sum\limits_{i=1}^q x^{u_i}_t$  

```{r define-enkf}
ensembleKalmanFilter <- function (y, X0, A, Q, C, R, Np) {
  N <- ncol(y)
  dimX <- length(X0)
  X <- array(X0,dim=c(dimX,Np),dimnames=list(vbl=colnames(C),particle=seq_len(Np)))
  filterMeans <- array(dim=c(dimX,N),dimnames=dimnames(x))
  predMeans <- array(dim=c(dimX,N),dimnames=dimnames(x))
  forecast <- array(dim=c(dimY,N),dimnames=dimnames(y))
  condlogLik <- numeric(N)
  tt <- c(0,t)
  for (k in seq_along(t)) {
    ## advance ensemble according to state process
    X[,] <- rprocess(pf,xstart=X,times=tt[c(k,k+1)],params=coef(pf),offset=1)
    predMeans[,k] <- pm <- rowMeans(X) # prediction mean
    ym <- C%*%pm                  # forecast mean
    vx <- var(t(X))               # prediction variance
    vy <- tcrossprod(C%*%vx,C)+R  # forecast variance
    vyx <- C%*%vx                 # forecast/state covariance
    Kt <- solve(vy,vyx)           # Kalman gain (transposed)
    Ek <- sqrtR %*% matrix(rnorm(n=dimY*Np),dimY,Np)
    X <- X+crossprod(Kt,y[,k]+Ek-C%*%X)
    resid <- y[,k]-ym
    svdS <- svd(vy,nv=0)
    condlogLik[k] <- sum(dnorm(x=crossprod(svdS$u,resid),mean=0,sd=sqrt(svdS$d),log=TRUE))
    filterMeans[,k] <- fm <- rowMeans(X)         # filter mean
    forecast[,k] <- ym
  }
  list(filterMeans=filterMeans,predMeans=predMeans,
       forecast=forecast,condlogLik=condlogLik)
}
```

```{r run-enkf}
enkf <- ensembleKalmanFilter(y,X0,A,Q,C,R,Np=100)
```

Plot the results:
```{r plot-enkf,warning=FALSE}
melt(list(obs=list(true=y,pred=enkf$forecast),
          states=list(true=x,filter=enkf$filterMeans,pred=enkf$predMeans))) %>%
  dcast(vbl+time~L2) %>%
  ggplot(aes(x=time))+
  geom_point(aes(y=true),color='black')+
  geom_point(aes(y=pred),color='red',alpha=0.5)+
  geom_line(aes(y=filter),color='blue',alpha=0.3)+
  facet_grid(vbl~.,scales='free_y')+
  theme(legend.position="top")
```

## Ensemble adjustment Kalman filter (EAKF)

Ensemble: $X_t\in \mathbb{R}^{m\times q}$  
Prediction mean: $M_t=\langle X \rangle$ (ensemble average).  
Prediction variance: $V_t=\langle\langle X \rangle\rangle$ (ensemble variance).  
SVD of prediction variance: $V = Q_{V}\,D_{V}\,Q_{V}^T$  
Another SVD: $U=D_V^{1/2}\,Q_V^T\,C^T\,R^{-1}\,C\,Q_V\,D_V^{1/2}=Q_U\,D_U\,Q_U^T$  
Adjustment: $B=Q_V\,D_V^{1/2}\,Q_U\,(I+D_U)^{-1/2}\,D_V^{-1/2}\,Q_V^T$  
Kalman gain: $K=B\,V\,B^T\,C^T\,R^{-1}$  
Filter mean: $m_t=M_t+K_t\,(y_t-C\,M_t)$  
Updated ensemble: $x_{t}=B\,(X_t-M_t\,\mathbb{1})+m_t\,\mathbb{1}$  

```{r define-eakf}
ensAdjKalmanFilter <- function (y, X0, A, Q, C, R, Np) {
  N <- ncol(y)
  dimX <- length(X0)
  ## X is the ensemble
  X <- array(X0,dim=c(dimX,Np),dimnames=list(vbl=colnames(C),particle=seq_len(Np)))
  filterMeans <- array(dim=c(dimX,N),dimnames=dimnames(x))
  predMeans <- array(dim=c(dimX,N),dimnames=dimnames(x))
  forecast <- array(dim=c(dimY,N),dimnames=dimnames(y))
  condlogLik <- numeric(N)
  ri <- solve(R)
  tt <- c(0,t)
  for (k in seq_along(t)) {
    # advance ensemble according to state process
    X[,] <- rprocess(pf,xstart=X,times=tt[c(k,k+1)],params=coef(pf),offset=1)
    predMeans[,k] <- pm <- rowMeans(X) # prediction mean
    X <- X - pm
    pv <- tcrossprod(X,X)/(ncol(X)-1)  # prediction variance
    svdV <- svd(pv,nv=0)
    resid <- y[,k]-C%*%pm              # forecast error
    w <- t(C%*%svdV$u)
    w <- crossprod(w,svdV$d*w)+R       # forecast variance
    svdW <- svd(w,nv=0)
    condlogLik[k] <- sum(dnorm(x=crossprod(svdW$u,resid),mean=0,sd=sqrt(svdW$d),log=TRUE))
    u <- sqrt(svdV$d)*t(C%*%svdV$u)
    u <- tcrossprod(u%*%ri,u)
    svdU <- svd(u,nv=0)
    ## adjustment
    b <- svdV$u%*%(sqrt(svdV$d)*svdU$u)%*%(1/sqrt(1+svdU$d)/sqrt(svdV$d)*t(svdV$u))
    K <- tcrossprod(b%*%pv,b)%*%crossprod(C,ri)   # Kalman gain
    filterMeans[,k] <- fm <- pm+K%*%resid         # filter mean
    forecast[,k] <- C %*% pm
    X[,] <- b%*%X+fm[,]
  }
  list(filterMeans=filterMeans,predMeans=predMeans,forecast=forecast,condlogLik=condlogLik)
}
```

```{r run-eakf}
eakf <- ensAdjKalmanFilter(y,X0,A,Q,C,R,Np=100)
```

Plot the results:
```{r plot-eakf,warning=FALSE}
melt(list(obs=list(true=y,pred=eakf$forecast),
          states=list(true=x,filter=eakf$filterMeans,pred=eakf$predMeans))) %>%
  dcast(vbl+time~L2) %>%
  ggplot(aes(x=time))+
  geom_point(aes(y=true),color='black')+
  geom_point(aes(y=pred),color='red',alpha=0.5)+
  geom_line(aes(y=filter),color='blue',alpha=0.3)+
  facet_grid(vbl~.,scales='free_y')+
  theme(legend.position="top")
```

## Comparison of likelihoods

```{r comparison3}
kable(c(enkf=sum(enkf$condlogLik),eakf=sum(eakf$condlogLik),kf=sum(kf$condlogLik),pf=logLik(pf)))
```

## Effect of ensemble size
```{r ensemblesize}
require(foreach)
require(doMC)
registerDoMC(30)
set.seed(181995286L,kind="L'Ecuyer")
foreach (Np=round(4^seq(3,7,by=1)),
         .combine=rbind,.inorder=FALSE,
         .options.multicore=list(set.seed=TRUE,preschedule=FALSE)) %:%
    foreach (rep=1:5,.combine=rbind) %dopar% 
    {
        enkf <- ensembleKalmanFilter(y,X0,A,Q,C,R,Np=Np)
        eakf <- ensAdjKalmanFilter  (y,X0,A,Q,C,R,Np=Np)
        pf1 <- pfilter(pf,Np=Np)
        data.frame(Np=Np,
                   rep=rep,
                   enkf=sum(enkf$condlogLik),
                   eakf=sum(eakf$condlogLik),
                   pf=logLik(pf1))
    } %>%
    melt(id=c("Np","rep"),value.name="logLik",variable.name="method") -> dat
```
```{r ensemblesize-plot}
dat %>% 
  ggplot(aes(x=Np,y=logLik,color=method))+
  geom_smooth()+scale_x_log10()+
  geom_hline(yintercept=sum(kf$condlogLik))+
  theme_bw()
```
